Economic Theory of Fertility Explains Modern Natural Selection

### 2020-12-01 why-negative-selection.Rmd
tweaks and fixes

@@

```{r plot-means-over-time, fig.height = 5.5, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals. Green lines show a significant linear increase over time (p < 0.05/33). Red lines show a significant decrease."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change)) + 
      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
        fatten = 1.25) +
      facet_wrap(~score_name, ncol = 5) +
      my_hline +
      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
      theme(
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

Figure \@ref(fig:plot-income-educ-level) plots selection effects for
each polygenic score, grouping respondents by age of completing
full-time education, and by household income. Effects are larger and
more significant for the lowest income category, and for the lowest
education category. Note that the overall effect is not a simple average
of the effect among the different subgroups, because polygenic scores
may also shift respondents between the subgroups. For example, a high
PGS for educational attainment may correlate with fewer children among
early school leavers, but may also increase the age at which a
respondent leaves school.

```{r plot-income-educ-level, fig.cap = "Selection effects by education and income", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}

@@

sexes_theme <-  my_theme(
                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
                  legend.position      = "bottom", 
                  legend.box.spacing   = unit(0, "lines"),
                  legend.box.margin    = margin(),
                  legend.box.just      = "left",
                  legend.justification = "left",
                  legend.text          = element_text(size = 7),
                  legend.title         = element_text(size = 8)
                )
sexes_guides <- guides(
                  # fill  = guide_legend(title.position = "top"),
                  # color = guide_legend(title.position = "top")
                )

@@

dfr %>% 
      mutate(
        Controls = fct_relevel(Controls, "None")
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        facet_grid(Correlation ~ Controls, labeller = label_both) + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "Correlation") +
        scale_colour_brewer(type = "qual", palette = 2) +
        theme(legend.position = "none")

@@

```{r plot-res-wt, fig.cap = "Selection effects: weighted regressions."}

n_regs <- length(score_names)
res_wt_combined %>% 
  mutate(
    Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                "Age/Qual/AFLB", "None"),
    term     = pretty_names(term),
    term     = fct_reorder(term, estimate, order_abs(1))
  ) %>% 
  rename(p = p.value) %>% 
  ggplot(aes(
    x = estimate, 
    y = term, 
    shape = Weights, 
    fill = Weights, 
    colour = p < 0.05/{{n_regs}}
  )) +
  geom_point(size = 1.5, stroke = 0.5) +
  my_vline +
  pval_scale() +
  scale_fill_manual(values = c(
    "None"              = "grey15", 
    "Geographical"      = "dodgerblue4",
    "Age/Qualification" = "steelblue2",
    "Age/Qual/AFLB"      = "darkseagreen4"
  )) +
  scale_shape_manual(values = c(
    "None" = "square filled", 
    "Geographical" = "circle filled",
    "Age/Qualification" = "circle filled", 
    "Age/Qual/AFLB" = "circle filled"
  ))

@@

Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
polygenic scores. For comparison, we show regressions on the respondents'
generation (*number of children*), excluding respondents with no
children.

>>>>

# nasty hack to make add_ashe_income work later
source("~/import-ukbb-data/import-ukbb-data.R")

@@

```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals. Green triangles show a significant linear increase over time (p < 0.05/33). Red squares show a significant decrease."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
        fatten = 1.25) +
      facet_wrap(~score_name, ncol = 5) +
      my_hline +
      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
      scale_shape_manual(values = c("-" = "square", "+" = "triangle", "o" = "circle open")) +
      theme(
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

Figure \@ref(fig:plot-income-educ-level) plots selection effects for
each polygenic score, grouping respondents by age of completing
full-time education, and by household income. Effects are larger and
more significant for the lowest income category, and for the lowest
education category. Note that the overall effect is not a simple average
of the effect among the different subgroups, because polygenic scores
may also shift respondents between the subgroups. For example, a high
PGS for educational attainment may predict having fewer children among
early school leavers, but it may also increase the age at which a
respondent leaves school.

```{r plot-income-educ-level, fig.cap = "Selection effects by education and income. Each point represents a single bivariate regression of number of children on a PGS", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}

@@

sexes_theme <-  my_theme(
                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
                  legend.position      = "bottom", 
                  legend.box.spacing   = unit(0, "lines"),
                  legend.box.margin    = margin(-5, 0, 0, -5),
                  legend.margin        = margin(0, 0, 0, 0),
                  legend.box.just      = "left",
                  legend.justification = "left",
                  legend.text          = element_text(size = 7),
                  legend.title         = element_text(size = 8)
                )
sexes_guides <- guides(
                  # fill  = guide_legend(title.position = "top"),
                  # color = guide_legend(title.position = "top")
                )

@@

dfr %>% 
      mutate(
        Controls = fct_relevel(Controls, "None")
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        facet_grid(Correlation ~ Controls, labeller = label_both) + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "Correlation") +
        scale_colour_brewer(type = "qual", palette = 2) +
        theme(
          legend.position = "none",
          panel.spacing = unit(14, "pt")
        )

@@

```{r plot-res-wt, fig.cap = "Selection effects: weighted regressions"}

n_regs <- length(score_names)
res_wt_combined %>% 
  mutate(
    Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                "Age/Qual/AFLB", "None"),
    term     = pretty_names(term),
    term     = fct_reorder(term, estimate, order_abs(1))
  ) %>% 
  rename(p = p.value) %>% 
  ggplot(aes(
    x = estimate, 
    y = term, 
    shape = Weights, 
    fill = Weights, 
    colour = p < 0.05/{{n_regs}}
  )) +
  geom_point(size = 1.5, stroke = 0.5) +
  my_vline +
  pval_scale() +
  scale_fill_manual(values = c(
    "None"              = "grey15", 
    "Geographical"      = "dodgerblue4",
    "Age/Qualification" = "steelblue2",
    "Age/Qual/AFLB"      = "darkseagreen4"
  )) +
  scale_shape_manual(values = c(
    "None" = "square filled", 
    "Geographical" = "circle filled",
    "Age/Qualification" = "circle filled", 
    "Age/Qual/AFLB" = "circle filled"
  )) + 
  labs(x = "Effect size", y = "")

@@

Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
polygenic scores. For comparison, we show regressions on the respondents'
generation (*number of children*), where respondents with no
children have been excluded.
